---
title: "Food Service Workers"
author: "Aaron Kessler"
date: "3/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(janitor)
library(readxl)
library(writexl)
library(tigris)
library(sf)
library(tmap)
library(tmaptools)

#load saved processed data files from step 00
states_allrecs <- readRDS("processed_data/states_allrecs.rds")
msa_allrecs <- readRDS("processed_data/msa_allrecs.rds")
national_allrecs <- readRDS("processed_data/national_allrecs.rds")


#limit to just the 50 states and DC
states_allrecs <- states_allrecs %>% 
  filter(!st %in% c("GU", "VI", "PR"))

msa_allrecs <- msa_allrecs %>% 
  filter(!prim_state %in% c("GU", "VI", "PR"))


```

*Data source: BLS Occupational Employment Statistics (2018)* 
*https://www.bls.gov/oes/tables.htm*  
  
## Filtering to just food prep and server worker occupations

A description of whom this category includes:
https://www.bls.gov/oes/current/oes350000.htm


```{r }

state_foodserv_all <- states_allrecs %>% 
  filter(str_starts(occ_code, "35-"))

```

Check to see if any our missing or redacted
```{r }
 
state_foodserv_all %>% 
  filter(is.na(tot_emp))

```

Just a handful. No worries then, we can work with that.    


Do individually listed items add up to the total major? Look at a grand totals and then a few individual states to see.
```{r}

state_foodserv_all %>% 
  group_by(occ_group) %>% 
  summarise(sum(tot_emp, na.rm = TRUE))


state_foodserv_all %>% 
  filter(st == "NJ") %>% 
  group_by(occ_group) %>% 
  summarise(sum(tot_emp, na.rm = TRUE))

state_foodserv_all %>% 
  filter(st == "CA") %>% 
  group_by(occ_group) %>% 
  summarise(sum(tot_emp, na.rm = TRUE))

state_foodserv_all %>% 
  filter(st == "MO") %>% 
  group_by(occ_group) %>% 
  summarise(sum(tot_emp, na.rm = TRUE))




```

This is pretty close! There could be some NAs among the detailed categories which could increase the spread for certain states, but this appears to be what's going on.  
  

### States by overall totals

To keep things simple to start, we'll just look at the overarching total from the major category record for each state. 

```{r}

state_foodserv_totalsonly <- state_foodserv_all %>% 
  filter(occ_code == "35-0000") %>% 
  select(-occ_title, -occ_group, -h_mean)

state_foodserv_totalsonly

```

#Total people employed overall? Let's double-check check to make sure we know what kind of workers are included here and who isn't...if there are any known quirks to be aware of.
```{r}
#total
state_foodserv_totalsonly %>% 
  summarise(sum(tot_emp))

```

Which states have the highest rate of employment in this sector per 1000?
```{r}

state_foodserv_totalsonly %>% 
  arrange(desc(jobs_1000))

```

This is interesting...as some of the states with the lowest populations actually have the highest percentage of people in this industry.  
  
What's the range in values?

```{r}

state_foodserv_totalsonly %>% 
  summarise(minvalue = min(jobs_1000),
            maxvalue = max(jobs_1000))

```




### Metropolitan area totals

Let's look at the same thing, the grand totals for all the food prep/service workers - but this time pulling from the Metro area (MSA) data.

```{r}

msa_foodserv_totalsonly <- msa_allrecs %>% 
  select(-occ_title, -occ_group, -h_mean) %>% 
  filter(occ_code == "35-0000")

msa_foodserv_totalsonly

```

Which metro areas have the highest rate of workers in the industry?
```{r}

msa_foodserv_totalsonly %>% 
  arrange(desc(jobs_1000))


```

Well it looks like my hometown area in South Jersey, near Atlantic City, is number 3. Well that's wonderful. :(  
  
What's the range in values for per 1000 figure here so we have a sense of it.

```{r}

msa_foodserv_totalsonly %>% 
  summarise(minvalue = min(jobs_1000),
            maxvalue = max(jobs_1000))


```

That's a pretty large difference between the worst metros and the best.  
Let's take a quick look at what the distribution looks like with a historgram.

```{r}

ggplot(msa_foodserv_totalsonly, aes(x = jobs_1000)) +
  geom_histogram() +
  geom_vline(aes(xintercept=mean(jobs_1000)),
            color="darkblue", linetype="dashed", size=1)


```

The plot above shows that metro areas all the way to the right - our worst-case metros - really do have an outsized vulnerability compared with the rest of the country. The mean (shown by the blue line above) is about 100 per 1000, or 1 out of every 10 workers.



### Combined Food Preparation and Serving Workers, Including Fast Food

A more specific group within the larger sector, let's take a look. (Need to investiate just what this does and doesn't include a bit more, but including here for now.)

By state:

```{r}

state_foodserv_all %>% 
  select(-occ_group, -h_mean) %>% 
  filter(occ_code == "35-3021") %>% 
  arrange(desc(jobs_1000))


```

By metro area:
```{r}

msa_allrecs %>% 
  select(-occ_group, -h_mean) %>% 
  filter(occ_code == "35-3021") %>% 
  arrange(desc(jobs_1000))

```



# Slicing out a specific type of worker

### Waiters/Waitresses

Clearly wait staff is going to be hit hard even if some establishments stay open for take-out and delivery. They won't need servers for that, so they're likely to still lose out.

Let's look at just wait staff at the state level

```{r}
state_foodserv_all %>% 
  select(-occ_group, -h_mean) %>% 
  filter(occ_code == "35-3031") %>% 
  arrange(desc(jobs_1000))
```

And the metro area level...
```{r}
msa_allrecs %>% 
  select(-occ_group, -h_mean) %>% 
  filter(occ_code == "35-3031") %>% 
  arrange(desc(jobs_1000))

```

Well that's not good again for South Jersey. Both Ocean City and Atlantic City both listed here.



### Dishwashers

By state:

```{r}

state_foodserv_all %>% 
  select(-occ_group, -h_mean) %>% 
  filter(occ_code == "35-9021") %>% 
  arrange(desc(jobs_1000))


```

By metro area:
```{r}

msa_allrecs %>% 
  select(-occ_group, -h_mean) %>% 
  filter(occ_code == "35-9021") %>% 
  arrange(desc(jobs_1000))

```



### Bartenders

By state:

```{r}

state_foodserv_all %>% 
  select(-occ_group, -h_mean) %>% 
  filter(occ_code == "35-3011") %>% 
  arrange(desc(jobs_1000))


```

By metro area:
```{r}

msa_allrecs %>% 
  select(-occ_group, -h_mean) %>% 
  filter(occ_code == "35-3011") %>% 
  arrange(desc(jobs_1000))

```

There's an irony here to Ocean City having the most bartenders given it's a dry city - there are no bars. But of course this is the MSA which extends around it and to the south.


# MAPS

#### A few maps for the heck of it 

Bring in geospatial boundaries of MSA via CBSA tigris download, and states as well
```{r}
cbsa_geo <- core_based_statistical_areas(cb = TRUE)
states_geo <- states(cb = TRUE)

names(cbsa_geo)
names(states_geo)
```

We'll try joining this to our MSA totals for the industry major category 
```{r}

msa_foodserv_totalsonly

geo_msa_foodserv_totalsonly <- geo_join(cbsa_geo, msa_foodserv_totalsonly, "GEOID", "area")

```

Map it
```{r}
tm_shape(geo_msa_foodserv_totalsonly) +
  tm_polygons("jobs_1000")


```

### State map of total industry

```{r}

state_foodserv_totalsonly

geo_state_foodserv_totalsonly <- geo_join(states_geo, state_foodserv_totalsonly, "GEOID", "area")

tm_shape(geo_state_foodserv_totalsonly) +
  tm_polygons("jobs_1000")

```


```{r}



```











