---
title: "Home Healthcare Aides"
author: "Aaron Kessler"
date: "4/8/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(janitor)
library(readxl)
library(writexl)
library(kableExtra)
library(formattable)
# library(tigris)
# library(sf)
# library(tmap)
# library(tmaptools)


#load saved processed data files from step 00
states_allrecs <- readRDS("processed_data/states_allrecs.rds")
msa_allrecs <- readRDS("processed_data/msa_allrecs.rds")
national_allrecs <- readRDS("processed_data/national_allrecs.rds")


#limit to just the 50 states and DC
states_allrecs <- states_allrecs %>% 
  filter(!st %in% c("GU", "VI", "PR"))

msa_allrecs <- msa_allrecs %>% 
  filter(!prim_state %in% c("GU", "VI", "PR"))


# pull out just home health aides (code 31-1011)
national_homehealthaides <- national_allrecs %>% 
  filter(occ_code == "31-1011")

states_homehealthaides <- states_allrecs %>% 
  filter(occ_code == "31-1011") %>% 
  select(state, tot_emp, jobs_1000)

msa_homehealthaides <- msa_allrecs %>% 
  filter(occ_code == "31-1011") %>% 
  select(area_name, tot_emp, jobs_1000)


```


The [Bureau of Labor Statistics](https://www.bls.gov/oes/tables.html) puts out extensive data on the workforce, including details down to the state and metro level. (The most recent localized data is from 2018, that's what we'll use here.)  
  
We'll use the occupation of "Home Health Aides" for this analysis. More information on the category can be found [here](https://www.bls.gov/ooh/healthcare/home-health-aides-and-personal-care-aides.htm).  
  
Nationally, about 800,000 people work as such aides. They generally make little money despite the importance of their jobs in keeping people alive - averaging about $12/hr as of 2018.

<br>

## State and local level workforces 

Now let's analyze the BLS data more granularly to explore how different states compare with each other.  
  
Since areas with the highest populations naturally tend to have more workers of any type, including those at restaurants and bars, we'll account for that by using a rate of "jobs per 1,000 workers" so it's more apples-to-apples. E.g. for each 1,000 workers in a state or metro area, how many are home health aides. 
  
### States

Which states have the highest rate of employment in this sector per 1000? (Shown below in the *jobs_1000* column).  

```{r, echo=FALSE}

states_homehealthaides %>%
  arrange(desc(jobs_1000)) %>% 
  head(15) %>% 
  kable(format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left") %>% 
  column_spec(3, background = "yellow")

```

What's the range in values we're talking about states having by that measure: 

```{r, echo=FALSE}

states_homehealthaides %>% 
  summarise(minvalue = min(jobs_1000),
            maxvalue = max(jobs_1000)) %>% 
  kable() %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")

```

That's a pretty large difference between the worst and the best.  
Let's take a quick look at what the distribution looks like with a historgram. 
  
As we can see below, the distribtuion skews left, meaning most states have very few home health aides as a percentage of their workforces, while a smaller number of state have significantly more.

```{r, echo=FALSE}

ggplot(states_homehealthaides, aes(x = jobs_1000)) +
  geom_histogram() +
  geom_vline(aes(xintercept=mean(jobs_1000)),
            color="darkblue", linetype="dashed", size=1)


```

Which states have the *least* home health aides as a share of their workforces?

```{r, echo=FALSE}

states_homehealthaides %>%
  arrange(jobs_1000) %>% 
  head(15) %>% 
  kable(format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left") %>% 
  column_spec(3, background = "yellow")

```


### Metro Areas

Now we'll do the same thing but for **Metropolitan Statistical Areas (MSAs)**...

Which metro areas have the highest rate of employment in this sector per 1000 workers? 

```{r, echo=FALSE}

msa_homehealthaides %>%
  arrange(desc(jobs_1000)) %>% 
  head(15) %>% 
  kable(format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left") %>% 
  column_spec(3, background = "yellow")

```

What's the range in values we're talking about states having by that measure: 

```{r, echo=FALSE}

msa_homehealthaides %>% 
  summarise(minvalue = min(jobs_1000, na.rm = TRUE),
            maxvalue = max(jobs_1000, na.rm = TRUE)) %>% 
  kable() %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")

```

Once again, that's a pretty large difference between the worst and the best.  

As we can see below, the distribtuion skews way left, meaning many areas have very few home health aides as a percentage of their workforces, while a smaller number have more.

```{r, echo=FALSE}

msa_homehealthaides %>% 
  filter(!is.na(jobs_1000)) %>% 
  ggplot(aes(x = jobs_1000)) +
    geom_histogram() +
    geom_vline(aes(xintercept=mean(jobs_1000)),
            color="darkblue", linetype="dashed", size=1)


```

Which metro areas have the *least* home health aides as a share of their workforces?

```{r, echo=FALSE}

msa_homehealthaides %>%
  arrange(jobs_1000) %>% 
  head(15) %>% 
  kable(format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left") %>% 
  column_spec(3, background = "yellow")

```




<br>

#### Saving to a file

Finally, we'll save the underlying data in a single Excel file for easy sharing, in case you want to explore the entire list of all states/metros included here.

```{r}

sheets <- list("states_homehealthaides" = states_homehealthaides,
              "msa_homehealthaides" = msa_homehealthaides,
              "national_homehealthaides" = national_homehealthaides)
  
write_xlsx(sheets, "output/exported_homehealthaides.xlsx")


```
